\textls[-0]{\thetitle}

### 2023-03-21 metatheory2.tex
add π1 π2 to MLTT syntax display
resolve some comments

\Citeauthor{debruijn1964} indices and explicit substitutions make details about
binders and substitutions explicit.
%
%The change of the syntax formulation is important to the simplicity of our
%proof, as we want to handle the proof about substitutions and terms
%uniformly---thus making substitution as part of the syntax (explicit
%substitution) will be helpful. 
\EDJ{Why comment out this sentence?}
%
The form \var{n} represents a variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

To simplify, $\SubstWeak{n}$ is a short hand for $\pi_1^n~"id"$ and $\var{n}$ is a short hand for $\pi_2~\pi_1^{n}$. Thus during meta-theoretic reasoning, we will only deal with $\pi_1$ and $\pi_2$.\YZ{$\pi_1$ and $\pi_2$ should be added to the syntax figure}\EDJreply{I added it, it is right after sub/dbj/shift.}

@@

Finally, we have \ruleref{sub/dbj/shift} defined using \ruleref{sub/wk} and \ruleref{tm/var}. This rule applying substitution $\gm$ on the earlier part of the context. We usually omit $A$ in the $\gm^{\uparrow A}$ because it can be inferred from the context.\YZ{The shift rule has bugs: premise says n, conclusion says 1.}\EDJreply{Thanks, fixed.}

@@

Universe levels address the size issue---it is unsound to have a set of all sets
(or a universe of all types)~\cite{hurkens1995simplification}\EDJ{Why remove this citation?}.
The level of a universe specifies ``how large'' that universe is.
%that is basically what universe level is measuring.

>>>>

\Citeauthor{debruijn1964} indices and explicit substitutions make details about
binders and substitutions explicit.
%
Using explicit substitutions obviates the need for special treatment of
substitutions in the proofs, as substitutions are part of the syntax.
%The change of the syntax formulation is important to the simplicity of our
%proof, as we want to handle the proof about substitutions and terms
%uniformly---thus making substitution as part of the syntax (explicit
%substitution) will be helpful. 
%
The form \var{n} represents a variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

To simplify, $\SubstWeak{n}$ is a short hand for $\pi_1^n~"id"$ and $\var{n}$ is a short hand for $\pi_2~\pi_1^{n}$. Thus during meta-theoretic reasoning, we will only deal with $\pi_1$ and $\pi_2$.

@@

Finally, we have \ruleref{sub/dbj/shift} defined using \ruleref{sub/wk} and \ruleref{tm/var}. This rule applying substitution $\gm$ on the earlier part of the context. We usually omit $A$ in the $\gm^{\uparrow A}$ because it can be inferred from the context.

@@

Universe levels address the size issue---it is unsound to have a set of all sets
(or a universe of all types)~\cite{hurkens1995simplification}.
The level of a universe specifies ``how large'' that universe is.
%that is basically what universe level is measuring.
